---
title: "Creating effective slides"
author: "Yue Jiang"
date: "Duke University"
header-includes: \usepackage{bm}
output:
  xaringan::moon_reader:
  mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
css: "sta440-slides.css"
logo: img/sta199-sticker-icon.png
lib_dir: libs/font-awesome
nature:
  highlightStyle: github
highlightLines: true
countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 3.75,
	fig.width = 6.25,
	message = FALSE,
	warning = FALSE
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
```

### Horrible presentations

You know them when you see them

- [https://www.slideshare.net/anilkr123/car-and-technology](Example 1)
- [https://www.slideshare.net/Kshivets/lung-cancer-surgery-4936542](Example 2)

---

### Software

- PowerPoint / Keynote (easy, WYSIWYG)
- LaTeX Beamer / R Markdown packages (clean, consistent, eventually straightforward)
- Random online things like Prezi (don't do this)

---

### Overall to-dos

- Be direct and clear in your messaging
- Keep things simple and clean - no clutter or walls of text!
- Be consistent in design and formatting (e.g., font sizes, alignment, etc.)
- Avoid redundancy (e.g., reading directly from slides)
- Keep things readable
- Use visual interest to highlight your message

---

### Activity 3

Activity 3 will essentially be what we are doing in class today and in lab on
Monday. You will create a ten-slide presentation explaining the results of the
Reutzler et al. study (provided on Sakai) and provide an accompanying oral
script for your presentation. Data from the study will also be provided, in 
case you would like to create visualizations or explore/model the data beyond 
what was presented in the paper itself.

---

### What to include in a presentation?

In deciding what to include per slide, try to distill your message to one main
idea (or even sentence!) per slide. So, if your presentation has ten slides, 
what are the ten most important sentences you want to get across? Are there
accompanying visuals that might help your point?

.question[
Design a five-slide presentation (not including title page, references, etc.) 
for the NEJM paper explaining the DIG trial, and upload the .pdf to Gradescope. 
On each page, highlight the main message your slide is trying to convey. We will
discuss the submissions at the end of class.
]