---
title: "Spatial data analysis (3)"
author: "Yue Jiang"
date: "Duke University"
header-includes: \usepackage{bm}
output:
  xaringan::moon_reader:
  mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
css: "sta440-slides.css"
logo: img/sta199-sticker-icon.png
lib_dir: libs/font-awesome
nature:
  highlightStyle: github
highlightLines: true
countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 3.75,
	fig.width = 6.25,
	message = FALSE,
	warning = FALSE
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(spdep)
library(spatialreg)
```


### Today's data

```{r, echo = F, dpi = 300, fig.height = 4.5, fig.align = "center"}
nc <- read_sf("data/census_2010_nc_tract.shp")
nc <- nc %>%
  filter(STCOFIPS == 37063) %>%
  select(TRACT, NAME, POP2010)
durham <- read_csv("data/clean.csv") %>%
  filter(!is.na(long) & !is.na(lat)) %>%
  mutate(long = long * -1) %>%
  st_as_sf(., coords = c("long", "lat"), 
           crs = 4269, 
           remove = F, 
           agr = "constant")
ggplot(nc) +
  geom_sf(aes(fill = POP2010)) +
  geom_point(data = durham, 
             mapping = aes(x = long, y = lat, col = yrs_lived),
             alpha = 0.4) +
  scale_color_gradient(low = "#fee8c8", high = "#7f0000") +
  theme_bw() +
  scale_x_continuous(breaks=seq(-79, -78.7, by = 0.1)) +
  labs(x = "Long.", y = "Lat", col = "Years", fill = "Pop.")
```

---

### Clustering of time in Durham?

```{r}
library(ape)
dists <- as.matrix(dist(cbind(durham$long, durham$lat))) + 0.01
dists <- 1/dists
diag(dists) <- 0
Moran.I(durham$yrs_lived, dists, na.rm = T)
```



---

### Today's data

```{r}
head(durham)
```

---

### Which census tract are we in?

```{r}
which_blocks <- st_intersects(durham, nc)
which_blocks
```

---

### Which census tract are we in?

```{r}
tracts <- nc[as.numeric(which_blocks), c("NAME")] %>%
  st_drop_geometry()
durham <- cbind(durham, tracts) 
head(durham)
```

---

### How many residents did we survey?

```{r}
resps <- as.data.frame(table(tracts)) %>%
  rename("NAME" = "tracts",
         "num_resp" = "Freq")
nc <- left_join(nc, resps, by = "NAME")
nc$num_resp[is.na(nc$num_resp)] <- 0
nc$prop_resp <- nc$num_resp/nc$POP2010*100
```

---

### How many residents did we survey?

```{r, echo = F, fig.align= "center", fig.height = 4.5, dpi = 300}
ggplot(nc) +
  geom_sf(aes(fill = num_resp)) +
  geom_point(data = durham, 
             mapping = aes(x = long, y = lat),
             alpha = 0.4, pch = 4) +
  scale_fill_gradient(low = "#fee8c8", high = "#7f0000") +
  theme_bw() +
  scale_x_continuous(breaks=seq(-79, -78.7, by = 0.1)) +
  labs(x = "Long.", y = "Lat", col = "Years", fill = "# Residents")
```

---

### How many residents did we survey?

```{r, echo = F, fig.align= "center", fig.height = 4.5, dpi = 300}
ggplot(nc) +
  geom_sf(aes(fill = prop_resp)) +
  geom_point(data = durham, 
             mapping = aes(x = long, y = lat),
             alpha = 0.4, pch = 4) +
  scale_fill_gradient(low = "#fee8c8", high = "#7f0000") +
  theme_bw() +
  scale_x_continuous(breaks=seq(-79, -78.7, by = 0.1)) +
  labs(x = "Long.", y = "Lat", col = "Years", fill = "% of Tract")
```

---

### A modeling aside

```{r}
summary(lm(yrs_lived ~ own + race, 
           data = durham))
```

---

### A modeling aside

```{r}
summary(lm(yrs_lived ~ own + race, 
           data = durham[durham$zip == 27703,]))
```

---

### A modeling aside

```{r}
summary(lm(yrs_lived ~ own + race, 
           data = durham[durham$zip == 27707,]))
```

---

### A modeling aside

```{r}
summary(lm(yrs_lived ~ own + race, 
           data = durham[durham$NAME == "Census Tract 21",]))
```

---

### A modeling aside

```{r}
summary(lm(yrs_lived ~ own + race, 
           data = durham[durham$NAME == "Census Tract 18.08",]))
```

---

### A modeling aside

```{r, echo = F, message = F, fig.width = 8, fig.height = 6, dpi = 300}
regfun <- function(x)  {
  dat <- durham[durham$NAME == x, ]
  m <- lm(yrs_lived ~ own + race, data=dat)
  coefficients(m)[2]
}

t_names <- unique(durham$NAME[durham$NAME %in% nc$NAME[nc$num_resp > 4]])

beta_own <- sapply(t_names, regfun)

library(stringr)
names(beta_own) <- str_sub(names(beta_own), end=-5)

dotchart(sort(beta_own), cex = 0.7, xlab = "Est. Coef.")
```

---

### A modeling aside

```{r}
summary(lm(yrs_lived ~ own + race, 
           data = durham))
```


---

### A modeling aside

```{r, eval = F}
beta_dat <- data.frame(as.data.frame(beta_own), names(beta_own))
names(beta_dat) <- c("beta", "NAME")

nc <- merge(nc, beta_dat, all.x = T)

ggplot(nc) +
  geom_sf(aes(fill = beta)) +
  geom_point(data = durham, 
             mapping = aes(x = long, y = lat),
             alpha = 0.4, pch = 4) +
  scale_fill_gradient2(low = "darkred", mid = "white", high = "darkblue",
                       midpoint = 0) +
  theme_bw() +
  scale_x_continuous(breaks=seq(-79, -78.7, by = 0.1)) +
  labs(x = "Long.", y = "Lat", fill = "Coef.")
```



---

### A modeling aside

```{r, echo = F, fig.align = "center", fig.height = 4.5, dpi = 300}
beta_dat <- data.frame(as.data.frame(beta_own), names(beta_own))
names(beta_dat) <- c("beta", "NAME")

nc <- merge(nc, beta_dat, all.x = T)

ggplot(nc) +
  geom_sf(aes(fill = beta)) +
  geom_point(data = durham, 
             mapping = aes(x = long, y = lat),
             alpha = 0.4, pch = 4) +
  scale_fill_gradient2(low = "darkred", mid = "white", high = "darkblue",
                       midpoint = 0) +
  theme_bw() +
  scale_x_continuous(breaks=seq(-79, -78.7, by = 0.1)) +
  labs(x = "Long.", y = "Lat", fill = "Coef.")
```

---

### A modeling aside

.question[
Using what we've seen for the tract-level analyses, how might we create a 
regression model for point-valued data?
]

---

### Local regression

For global models so far, we've assumed that all observations in the dataset
are equally important in estimating regression coefficients. However, it might
be the case that nearer observations are more "important" than ones farther
away.

In local regression, this is precisely the case, with a weighting scheme based
on distance from each point. The relative importance of near vs. far points is
given by the .vocab[bandwidth] used in the .vocab[kernel] - a smaller bandwidth
is "more local" whereas a larger bandwidth is "more global." 

---

### Local regression

In local regression, we perform many separate regression models estimated at
reression points. In actuality, instead of a single point estimate of each
regression parameter $\beta$, we have a *surface* of such estimates (that are
estimated only at the observed points).

We note that there may be different local estimates of $\beta$ depending on 
where the point lies - this indicates the changing relationship(s) between
predictors and outcomes through space, which can be visualized. Note that we
might also map the SEs associated with each parameter estimate.

---

### Local regression

```{r}
library(spgwr)
spdf <- SpatialPointsDataFrame(
  coords = as.data.frame(st_drop_geometry(durham[,c("long", "lat")])), 
  data = as.data.frame(st_drop_geometry(durham)),
  proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))

band <- gwr.sel(yrs_lived ~ own + race, data = spdf, gweight = gwr.Gauss, verbose = TRUE)

m1 <- gwr(yrs_lived ~ own + race, data = spdf, bandwidth = band, gweight = gwr.Gauss, hatmatrix = T)
```

Also check out `ggwr` and `ggwr.sel` for generalized geographically weighted
regression.

---

### Local regression

```{r}
m1
head(m1$SDF$own)
```

---

### A modeling aside

```{r}
summary(lm(yrs_lived ~ own + race, 
           data = durham))
```

---

### Local regression

```{r, eval = F}
durham <- cbind(durham, m1$SDF$own)

ggplot(nc) +
  geom_sf(fill = "darkgray") +
  geom_point(data = durham, 
             mapping = aes(x = long, y = lat, col = m1.SDF.own),
             alpha = 0.5, pch = 19) +
  scale_color_gradient2(low = "darkred", mid = "white", high = "darkblue",
                       midpoint = 0) +
  scale_fill_gradient2(low = "darkred", mid = "white", high = "darkblue",
                       midpoint = 0) +
  theme_bw() +
  scale_x_continuous(breaks=seq(-79, -78.7, by = 0.1)) +
  labs(x = "Long.", y = "Lat", col = "Coef.")
```


---

### Local regression

```{r, echo = F, fig.align = "center", fig.height = 4.5, dpi = 300}
durham <- cbind(durham, m1$SDF$own)

ggplot(nc) +
  geom_sf(fill = "darkgray") +
  geom_point(data = durham, 
             mapping = aes(x = long, y = lat, col = m1.SDF.own),
             alpha = 0.5, pch = 19) +
  scale_color_gradient2(low = "darkred", mid = "white", high = "darkblue",
                       midpoint = 0) +
  scale_fill_gradient2(low = "darkred", mid = "white", high = "darkblue",
                       midpoint = 0) +
  theme_bw() +
  scale_x_continuous(breaks=seq(-79, -78.7, by = 0.1)) +
  labs(x = "Long.", y = "Lat", col = "Coef.")
```
